---
insttitle: "Project Deliverable 1"
author: "Rees Klintworth"
date: "September 22, 2014"
output: word_document
---

```{r}
# Functions that will be used by the rest of my code
analyze_timeseries <- function(ts, symbol)
  {
    h_value = round(length(ts)/10)
    
    # Holt-Winters with both parameters true
    beta_gamma_forecasts <- HoltWinters(ts)
    beta_gamma_forecasts2 <- forecast.HoltWinters(beta_gamma_forecasts, h=h_value)
    beta_gamma_box <- Box.test(beta_gamma_forecasts2$residuals, lag=h_value, type="Ljung-Box")
  
    # Holt-Winters with both parameters false
    neither_forecasts <- HoltWinters(ts, beta=FALSE, gamma=FALSE)
    neither_forecasts2 <- forecast.HoltWinters(neither_forecasts, h=h_value)
    neither_box <- Box.test(neither_forecasts2$residuals, lag=h_value, type="Ljung-Box")
  
    # Holt-Winters with gamma false
    beta_forecasts <- HoltWinters(ts, gamma=FALSE)
    beta_forecasts2 <- forecast.HoltWinters(beta_forecasts, h=h_value)   
    beta_box <- Box.test(beta_forecasts2$residuals, lag=h_value, type="Ljung-Box")
    
    # Checks for all forecasts where the p value is greater than the threshold (0.1)
    forecasts <- c(beta_gamma_forecasts2, neither_forecasts2, beta_forecasts2)
    p_values <- c(beta_gamma_box$p.value, neither_box$p.value, beta_box$p.value)
    matching_models <- p_values >= 0.1
    
    matched = FALSE
    
   if (matching_models[1])
     {
      plot.forecast(beta_gamma_forecasts2, main="")
      title(paste0(symbol, " Holt-Winters: Beta, Gamma"))
      cat("Holt-Winters with Beta and Gamma fit well with a p-value of", p_values[1], "\n")
      matched = TRUE
     }
   
   if (matching_models[2])
     {
      plot.forecast(neither_forecasts2, main="")
      title(paste0(symbol, " Holt-Winters"))
      cat("Holt-Winters without Beta or Gamma fit well with a p-value of", p_values[2], "\n")
      matched = TRUE
     }
   
   if (matching_models[3])
     {
      plot.forecast(beta_forecasts2, main="")
      title(paste0(symbol, " Holt-Winters: Beta"))
      cat("Holt-Winters with Beta fit well with a p-value of", p_values[3], "\n")
      matched = TRUE
     }
   
   if (!matched)
     {
      cat("No model fit the data well", "\n")
     }
  }

# This function calculates two moving averages based on inputed day values.
# Data must have an "Adjusted Close" column
calculate_moving_averages <- function(data, days1, days2, symbol)
  {
    # Find the first moving average
    data_sma1 <- SMA(data$`Adjusted Close`, n = days1)
    # Find the second moving average
    data_sma2 <- SMA(data$`Adjusted Close`, n = days2)
    # Approximate Frequency by finding average number of entries per year
    number_per_year <- round(nrow(data)/13)
    # Create a data frame from the three columns of data
    data_frame <- data.frame(data$`Adjusted Close`, data_sma1, data_sma2)
    # Create a time series from the data frame
    time_series <- ts(data_frame, freq=number_per_year, start=c(2001,1))
    colors <- c("Black", "Red", "Blue")
    # Output the timeseries
    ts.plot(time_series, col=colors, xlab="Year", ylab="Price (American Dollars)")
    title(paste0(symbol, " Adjusted Close and Moving Averages"))
    sma1_text <- paste0(days1, " Days")
    sma2_text <- paste0(days2, " Days")
    legend(x=2002, y=max(data$`Adjusted Close`), legend=c("Orignal", sma1_text, sma2_text), lwd=1, col = colors)
  }
```

1. For the first part of this assignment, I have read in VFINX and TNX data from Yahoo! via Quandl. Below, I will display the structure (str function) and first 10 items (head function) of each dataset.
```{r, comment=NA}
# Load necessary packages
suppressMessages(library(Quandl))
suppressMessages(library(TTR))
suppressMessages(library(forecast))
Quandl.auth("Xwpyys22sxHPzyXBrGdH")

# US stock market values from 1980
vfinx <- Quandl("YAHOO/FUND_VFINX")
cat("VFINX data:", "\n")
head(vfinx)
str(vfinx)
# Risk-free rate of return
tnx <- Quandl("YAHOO/INDEX_TNX")
cat("TNX data:", "\n")
head(tnx)
str(vfinx)

```
2. At this stage in the project, data cleaning was minimal because it wasn't necessary. I took necessary steps throughout coding to prevent errors, and will be able to easily add more functionality later.

3. I picked ten different stocks (with no rhyme or reason) to analyze. The companies represented are: Apple, Adobe, Best Buy, Deere & Company, eBay, Garmin, Coca-Cola, Pepsi, Under Armour, and Union Pacific.

I gathered data for each stock from January 1, 2001 to December 31, 2013. This can be changed easily in the future for more responsive data analysis. I also attempted to fit each stock to a Holt-Winters model, experimenting with different values of Beta and Gamma. If any model fit with a p-value greater than or equal to .1, I displayed the output. Note that, because of my large time range, there were very few good fits. The code is established however, and would yield more printed results for a narrower time period.

4. I used the diff function and simple mathematics to calculate the rate of return for each stock.

5. I plotted the Adjusted Closing prices for each stock, along with two moving averages (one large and one small).

The output from parts 3 - 5 can be seen below for each stock.

```{r, comment=NA}
# These are the 10 stocks I selected to analyze
my_stocks <- c("AAPL", "ADBE", "BBY", "DE", "EBAY", "GRMN", "KO", "PEP", "UA", "UNP")

# Loop through stock symbols
invisible(sapply(my_stocks, function(symbol)
  {
    cat("Analyzing", symbol, "\n")
    # Create Quandl code
    quandl_code = paste0("YAHOO/", symbol)
    # All of the stocks go back to at least the beginning of 2001, and wanted analyze the most years possible
    my_data <- Quandl(quandl_code, start_date="2001-01-01", end_date="2013-12-31")
    # Reverse the data so it isn't backwards
    my_data <- my_data[nrow(my_data):1,]
    # Determine frequency
    number_per_year <- round(nrow(my_data)/13)
    # Create timeseries
    my_ts <- ts(my_data[,'Adjusted Close'], freq=number_per_year, start=c(2001,1))
    analyze_timeseries(my_ts, symbol)
    # Find the return rates, and multiply by number per year for an annualized value
    return_rates <- diff(my_data$`Adjusted Close`)
    annualized_return_rate <- mean(return_rates) * number_per_year
    hist(return_rates)
    cat("Annual return rate:", annualized_return_rate, "\n") 
    calculate_moving_averages(my_data, 50, 200, symbol)
  }))
```